Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-08 Manuscript/manuscript.Rmd
address ???

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by PEAKACHU at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. [??? Can you look up closer? If Arrowhead could call boundaries at 5kb, but not 10kb, it is strange; !!!https://groups.google.com/forum/#!searchin/3d-genomics/chr9$20from$20k562%7Csort:date]. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2). [??? Change "cis-regulatory elements (CRE)" to "C; !!!fixed]

@@

Cell type-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average [??? Is that correct? Or, max?; !!!yes average] signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes. [??? Of what size?; !!!not sure what question means]

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *P*, to consider at each node. The number of trees that were aggregated for each RF model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity: [??? Correct?;!!!yep]

@@

Additionally, we compared the $log_2$ [??? log2?;!!!added] distance between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test. The alpha level of significance was set to 0.05.

Furthermore, under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell types and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) [??? Package?;!!!added] and evaluated using the Jaccard index defined as

@@

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview] [??? Read 10.1002/widm.1072 and 10.1002/widm.1301, elaborate, cite;!!!added first source]. We then leveraged our optimally built RF to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called [??? Did you call them?; !!!yes] using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions, and these results were consistent for loop boundaries (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects model performance [@prati2011survey; @wei2013role] [??? Add refs; !!!added]. To address class imbalance, we evaluated the effect of three re-sampling techniques on prediction performance. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote] (see Methods). These strategies allowed us to select the best performing model.

>>>>

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by PEAKACHU at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2).

@@

Cell type-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *P*, to consider at each node. The number of trees that were aggregated for each RF model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

Additionally, we compared the $log_2$ distance between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test. The alpha level of significance was set to 0.05.

Furthermore, under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell types and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview]. We then leveraged our optimally built RF to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions, and these results were consistent for loop boundaries (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects model performance [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques on prediction performance. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote] (see Methods). These strategies allowed us to select the best performing model.
